Implementing a Generic Radix Trie in Rust

### 2015-03-15 Paper.tex
CMPS 112 version.

\documentclass[a4paper,12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumerate}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{minted}
\usepackage[T1]{fontenc}
\setlength\parindent{0pt}
\setcounter{secnumdepth}{0}
\pagestyle{empty}

@@

Rust is a systems programming language developed by Mozilla Research that is rapidly approaching its 1.0 release. Built on an imperative core with C-like syntax, Rust also includes several sophisticated features inspired by the ML family of languages. Notable features include type generics, traits (similar to Haskell's type classes), type inference, algebraic data types and pattern matching.  Rust's status as a systems programming language is established by its memory management system, which guarantees memory safety without garbage collection. It achieves this through a system of ownership and borrowing inspired by the Cyclone Programming Language [reference].\\

@@

Rust's ownership system [reference] interacts in several interesting ways with hierarchical data-structures. The following two sections discuss my experiences with Rust's memory management, including some ownership related challenges and their solutions.

@@

One negative side-effect of the delayed action approach is that it prevents the function from being tail-recursive, thus destroying the opportunity to perform tail call optimisation. Within the context of Rust this turns out not to be a problem, as tail call optimisation was deemed to interact badly with Rust's memory model and is unlikely to be added in the near future [reference].

@@

Rust's strict distinction between mutable and immutable values becomes slightly awkward when writing functions that differ only in the mutability of their arguments or results. Following the design of the Rust standard data-structures I set out to provide both \code{get} and \code{get\_mut} methods for my radix trie. The logic for these two methods is identical except for the sprinkling of a few \code{mut} keywords, which present-day Rust doesn't provide syntax to abstract over. Several Rust developers have suggested a syntax similar to the following [reference]:

@@

To work around the lack of parametric mutability I turned to Rust's macro system, which was recently stabilised in preparation for Rust 1.0 [reference]. Although generally agreed to be less than ideal, the macro system includes several nice features including hygiene and separator-based capturing of multiple items. As the two get functions differ only in the placement of the \code{mut} keyword, I designed the macro to take the keyword (or lack of keyword) as an argument, drawing inspiration from the approach used in TrieMap [reference]. The resulting macro signature looks like this:

@@

@@

@@

@@

https://github.com/rust-lang/rust/issues/217

https://github.com/michaelsproul/rust\_nibble\_vec/commit/69edc98774656ec0b44ede9b34fce0d4ddb57fbb

https://github.com/rust-lang/rust/commit/a4d22635e1453d77e90c4335f5c2e1f62e3185eb

http://internals.rust-lang.org/t/parameterisation-over-mutability/235

https://github.com/rust-lang/rfcs/blob/master/text/0453-macro-reform.md

https://github.com/Gankro/collect-rs/blob/8392f925094a947629327195cce405bfdf11ba0e/src/trie/map.rs\#L435

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{enumerate}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{minted}
\usepackage[T1]{fontenc}
\setlength\parindent{0pt}
\setcounter{secnumdepth}{0}
\pagestyle{empty}

@@

Rust is a systems programming language developed by Mozilla Research that is rapidly approaching its 1.0 release. Built on an imperative core with C-like syntax, Rust also includes several sophisticated features inspired by the ML family of languages. Notable features include type generics, traits (similar to Haskell's type classes), type inference, algebraic data types and pattern matching.  Rust's status as a systems programming language is established by its memory management system, which guarantees memory safety without garbage collection. It achieves this through a system of ownership and borrowing inspired by the Cyclone Programming Language.\\

@@

Rust's ownership system interacts in several interesting ways with hierarchical data-structures. The following two sections discuss my experiences with Rust's memory management, including some ownership related challenges and their solutions.

@@

One negative side-effect of the delayed action approach is that it prevents the function from being tail-recursive, thus destroying the opportunity to perform tail call optimisation. Within the context of Rust this turns out not to be a problem, as tail call optimisation was deemed to interact badly with Rust's memory model and is unlikely to be added in the near future.

@@

Rust's strict distinction between mutable and immutable values becomes slightly awkward when writing functions that differ only in the mutability of their arguments or results. Following the design of the Rust standard data-structures I set out to provide both \code{get} and \code{get\_mut} methods for my radix trie. The logic for these two methods is identical except for the sprinkling of a few \code{mut} keywords, which present-day Rust doesn't provide syntax to abstract over. Several Rust developers have suggested a syntax similar to the following:

@@

To work around the lack of parametric mutability I turned to Rust's macro system, which was recently stabilised in preparation for Rust 1.0. Although generally agreed to be less than ideal, the macro system includes several nice features including hygiene and separator-based capturing of multiple items. As the two get functions differ only in the placement of the \code{mut} keyword, I designed the macro to take the keyword (or lack of keyword) as an argument, drawing inspiration from the approach used in TrieMap. The resulting macro signature looks like this:

@@

To test the performance of my implementation I benchmarked it against open source implementations written in Go and Haskell.

\begin{itemize}
\item \textbf{Haskell:} bytestring-trie 0.2.4, \url{https://hackage.haskell.org/package/bytestring-trie}
\item \textbf{Go:} go-patricia 2.1.0, \url{https://github.com/tchap/go-patricia}
\end{itemize}

Approximately equivalent benchmarks were written in each language for each of the three main operations; get, insert and remove. Performance metrics were then recorded using two texts as input: \textit{The Sun Rising} by John Donne and \textit{Nineteen Eighty-Four} by George Orwell.\\

\begin{table}
\begin{center}
\begin{tabular}[h]{| l | l | l | l |}
\hline
                & \textbf{Rust} & \textbf{Haskell}  & \textbf{Go}   \\ \hline
get             & 0.113         & 0.115             & 0.061         \\ \hline
insert          & 0.143         & 0.067             & 0.284         \\ \hline
insert-remove   & 0.280         & -                 & 0.350         \\ \hline
remove          & 0.137*        & 0.110             & 0.066*        \\ \hline
\end{tabular}
\caption{Results for \textit{The Sun Rising} (226 words); times are in milliseconds.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}[h]{| l | l | l | l |}
\hline
                & \textbf{Rust} & \textbf{Haskell}  & \textbf{Go}   \\ \hline
get             & 88            &  55               & 40            \\ \hline
insert          & 93            &  409              & 57            \\ \hline
insert-remove   & 175           &  -                & 97            \\ \hline
remove          & 82*           &  387              & 38*           \\ \hline
\end{tabular}
\caption{Results for \textit{Nineteen Eighty-Four} (103,366 words); times are in milliseconds.}
\end{center}
\end{table}

The insert-remove benchmark for Go and Rust records the time to insert then remove each word of the entire text. This was done because both the Go and Rust tries lack deep-clone implementations. The starred (*) values indicate ones estimated by subtracting the insert benchmark from the remove benchmark (pre-rounding).\\

Although the Haskell library used for benchmarking (Criterion) outputs statistical information, no attempt has been made to determine statistical significance in the Rust and Go benchmarks. For this reason, these numbers serve only as rough estimates of performance. Most importantly these benchmarks indicate that my trie implementation achieves performance close to that of existing implementations. The Rust radix trie comes in second quite consistently, and with some further performance tweaks I'm confident it could challenge the Go implementation. Profiling revealed a lot of time spent in calls to \code{NibbleVec::split}, which could be reduced by using \textit{slicing} instead of copying when recursing.

@@

There are few more things to do before the library is ready for consumption by the wider Rust community, but I'm confident I'll be able to accomplish these things in the coming months.

\begin{itemize}
\item Implement the ``collection views" API, to allow efficient updating after node location.
\item Improve performance by writing a slicing method for \code{NibbleVec}.
\item Provide \code{TrieKey} implementations for more in-built Rust types (macros will probably be useful for the numeric types).
\item Publish the code on \url{crates.io}!
\end{itemize}

@@

Rust proved to be a very pleasant language for implementing data structures. Although some features of the language required conceptual re-adjustments, the safety guarantees provided by the ownership system are worth the effort. As one of very few modern languages without garbage collection I hope to see Rust give C and C++ a run for their money. Rust's modern language features in combination with its robust build system should see it succeed even outside the systems programming space. Although the lack of parametric mutability makes some situations slightly awkward, for a language that hasn't reached 1.0 yet, I have very few complaints about Rust.

@@

\textit{PATRICIA - Practical Algorithm To Retrieve Information Coded in Alphanumeric}, Donald Morrison, 1968.

\url{https://github.com/rust-lang/rust/issues/217}

\url{https://github.com/michaelsproul/rust\_nibble\_vec/commit/69edc98774656ec0b44ede9b34fce0d4ddb57fbb}

\url{https://github.com/rust-lang/rust/commit/a4d22635e1453d77e90c4335f5c2e1f62e3185eb}

\url{http://internals.rust-lang.org/t/parameterisation-over-mutability/235}

\url{https://github.com/rust-lang/rfcs/blob/master/text/0453-macro-reform.md}

\url{https://github.com/Gankro/collect-rs/blob/8392f925094a947629327195cce405bfdf11ba0e/src/trie/map.rs\#L435}
